---
title       : Is that Mushroom Poisonous or Edible?
subtitle    : Classifying mushrooms by descriptive features
author      : Ty Abonil
job         : 
framework   : io2012        # {io2012, html5slides, shower, x`dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---
### The Problem

The goal is to create a web app that ascertains whether a given mushroom is edible or poisonous, given qualitative description of its physical characteristics.

[The Mushroom Data Set (MDS)](https://archive.ics.uci.edu/ml/datasets/Mushroom) is a multivariate collection of 8124 instances of mushroom descriptions, each of 22 discrete attributes, including a classification of whether a mushroom is edible or poisonous.  22 features is daunting.
-  Bache, K. & Lichman, M. (2013). [UCI Machine Learning Repository](http://archive.ics.uci.edu/ml). Irvine, CA: University of California, School of Information and Computer Science.


To tackle the problem:
- The features must be reduce without impacting accuracy of classification
- The model must be very accurate as an incorrect classifcation can be deadly


```{r, echo=FALSE}
library(caret)
library(nnet)
library(rpart)
library(rattle)
```
The file is read and, using the schema provided on the UCI website, names are assigned. NA values are omitted and 50/50 test and training sets are created. Please see the source file for details. 


```{r, echo=FALSE}
set.seed(1)
mush.raw.orig <- read.csv("../mushrooms/agaricus-lepiota.data", header=F, na.string="?")
names(mush.raw.orig) <- c("edible", "cap.shape", "cap.surface", "cap.color", "bruises", "odor", "gill.attachment", "gill.spacing", "gill.size", "gill.color", "stalk.shape", "stalk.root", "stalk.surface.above.ring", "stalk.surface.below.ring", "stalk.surface.below.ring", "stalk.color.below.ring", "veil.type", "veil.color", "ring.number", "ring.type", "spore.print.color", "population", "habitat")
mush.raw.clean  <-  na.omit(mush.raw.orig)
mush.raw.clean <- mush.raw.clean[, !(names(mush.raw.clean) %in% c("REPEAT"))]
mush.raw.clean <- mush.raw.clean[sample(nrow(mush.raw.clean)),] ## shuffle the dataframe 

## create training and testing data sets
mush.train_idx  <- sample(1:nrow(mush.raw.clean), nrow(mush.raw.clean)/2, replace=FALSE)
mush.train  <- mush.raw.clean[mush.train_idx, ]
mush.test  <- mush.raw.clean[-mush.train_idx, ]
write.csv(mush.train, "mush.train.csv")
write.csv(mush.test, "mush.test.csv")
```

--- &twocol

### Reducing Featureset
We opt to use the rpart package to grow a decision tree to find the most valuable features.  First, we create a tree with all possible features, then we prune the tree to find the ones which minimize the CP hyper parameter - a measure of the 100-fold cross-validated error.

*** {name: left}


```{r, echo=FALSE}
mush.frm = edible ~ .
mush.fit <- rpart(mush.frm, method="class", data=mush.train,
                  control=rpart.control(minsplit = 2,
                                        cp = 0.01,
                                        xval = 100
                  )
)
mush.pfit<- prune(mush.fit, cp=   mush.fit$cptable[which.min(mush.fit$cptable[,"xerror"]),"CP"])
fancyRpartPlot(mush.fit, uniform=TRUE, 
    main="Classification Tree")
text(mush.fit, use.n=TRUE, all=TRUE, cex=.8)
```

*** {name: right}

```{r, echo=FALSE}
fancyRpartPlot(mush.pfit, uniform=TRUE, 
    main="Pruned Classification Tree")
text(mush.pfit, use.n=TRUE, all=TRUE, cex=.8)
```

It is osberved that pruning down to two features has no effect on the tree.

---

### The ANN Model


It is observed that "odor" and "spore.print.color" account for almost all of the variance in the final classification.  An ANN is trained using these values:

```{r trainChunk, cache=TRUE}

mush.frm <- edible ~ odor + spore.print.color
mush.trained <- train(as.formula(mush.frm), data=mush.train, method = "nnet", 
                      maxit = 100,  trace = F)
```

We see that the best ANN is at size ```r mush.trained$bestTune$size``` and decay ```r mush.trained$bestTune$decay```.

```{r}
mush.fit <- nnet(as.formula(mush.frm), data=mush.train, method = "nnet", 
                 maxit = 100,  trace = F, size=1, decay=0.1)
```

---

### The Validation Test

Using the validation set, the accuracy of the model is tested:

```{r}
success.rate <- sum(
  predict(
    mush.fit, newdata=mush.test, type="class") == mush.test[,1]
  )/nrow(mush.test) * 100
success.rate
```

The success rate is found to be `r success.rate`%!  This is a very strong model, so we can use it to predict whether a mushroom is poisonous or not.
